Cryptanalysis of post-quantum signature scheme while breaking dlog

### 2025-05-20 rapport.tex
version envoyé à Sorina

\author{Paul Mekhail\\ \textbf{Encadrante:} Sorina Ionica}

@@

		\begin{definition}
			Let $\prec$ be an admissible monomial ordering on a given ring $R = \mathbb{K}[x_1,\dots,x_n]$ where $\mathbb{K}$ is a field. Given a sequence of homogeneous (resp. affine) polynomials $\mathcal{F} = (f_1,\dots,f_m) \in R$, we associate to it the Macaulay matrix of degree $D$ (resp. $\leq D$), denoted by $\Mac{D}{m}{\mathcal{F}}$ (resp. $\Mac{\leq D}{m}{\mathcal{F}}$), and defined as follows: the columns of the matrix are indexed by the monomials of degree $D$ (resp. $\leq D$) sorted in decreasing order from the left to right using the defined monomial ordering. Each row of the matrix is labeled by a tag, also called signature, ($u, f_i$) where u is a monomial in $R$ and $f_i \in \mathcal{F}$ such that $deg(uf_i) = D$ (resp. $deg(uf_i) \leq D$), and the polynomial $uf_i$ is written as vector of coefficients of monomials on the row. We denote by $\Mact{D}{m}{F}$ (resp. $\Mact{\leq D}{m}{F}$) the row echelon form of $\Mac{D}{m}{\mathcal{F}}$ (resp. $\Mac{\leq D}{m}{\mathcal{F}}$) without swapping columns to preserve the monomial ordering.
		\end{definition}

@@

		The major "issue" of Gröbner basis algorithms is the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
		reductions to 0 can cost a lot during the linear algebra process. F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.

@@

		\subsection{Hilbert series}
		Hilbert-Poincaré series (or just Hilbert series) is a very important tool in algebraic geometry to describe ideals and varieties and their dimensions.
		\begin{definition}[\cite{CLS}]
			Let $I$ be a homogeneous ideal in $R$. The affine Hilbert function of $I$ is the function on the nonnegative integers s defined by
			$$
				{}^a HF_{R/I}(s) = dim R_{s} / dim I_{s} = dim R_{s} - dim I_{s}.
			$$
		\end{definition}

@@

		Thus, Bardet came with a new concept called semi-regularity for overdetermined system to allow their study and analysis. The definition of a semi-regular sequence is given by Bardet \cite{Bardet04}.

@@

		\subsection{Boolean systems}

		The problem of finding solutions to a multivariate polynomial system over a finite field $\mathbb{F}_p$ is proven NP-complete in \cite{FY79}. It is also well known that solving a quadratic polynomial system over any finite-field is NP-hard. This does not mean that any random instance is hard to solve. The problem of the MQ problem is that some instances can be very easy to solve, so the goal of cryptographers is to imagine a system with parameters that can be hard to solve in general and not have easy cases to solve most of the time.

		Sometimes, the finite field used is $\mathbb{F}_2$ because it is still supposedly hard to solve but is very easy to use and manipulate on computers. But in such small fields one can begin to use exhaustive search to obtain the solution to the system. The best algorithm known to the authors for exhaustive search for multivariate polynomial system $f_1,\dots,f_m$ over $\mathbb{F}_2[x_1,\dots,x_n]$ is \cite{BCC+10} with a complexity $\mathcal{O}(ln(n)2^n)$, what is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$. There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, libFES.

		The last type is a hybrid algorithms using algebraic geometry and gröbner basis algorithms and exhaustive search to find solutions. The most known and the best one is BooleanSolve hybrid by Bardet, Faugère, Salvy and Spaenlehauer \cite{BFSS13} where we guess a certain number of variables $k$ and apply F5 algorithm on the specialized systems. Under semi-regularity assumption, the algorithm finds the solution in $\mathcal{O}^(2^{0.792n})$ if $n = m$ and $k = 0.55n$. This is better than exhaustive search, but still, the constant factors are too big for it to be better for systems smaller with less than 200 variables.

		\begin{algorithm}
			\caption{The CrossBred algorithm}\label{alg:crossbred}
			\begin{algorithmic}[1]
				\Require Polynomials $\mathcal{F} = {f_1, \dots, f_m}$ of $n$ variables and integers $D,d,k$
				\Ensure  A solution for the system if it exists or nothing otherwise.
				\State Construct $\Mac{arg1}{arg2}{arg3}$
				\State Compute a basis $(v_1,\dots,v_r)$ if the left kernel of ...
				\State Construct the polynomials $p_1,\dots,p_r$ corresponding ti ...
				\For{$a = (a_{k+1},\dots,a_n) \in \mathbb{F}_{2}^{n-k}$}
					\State a
					\State a
					\State a
					\State a
					\If{$S*$ is consistent using XL for example}
						\State \textbf{Return} solution
					\EndIf
				\EndFor
			\end{algorithmic}
		\end{algorithm}

@@

		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$ which is the ideal generated by the polynomials of the system.

@@

		When applying Macaulay matrix generation code from hpXbred \cite{BS23} on NSBC polynomial systems, we notice that they are not semi-regular when we only multiply by the $x$ variables or only by the $y$ variables. If we multiply by the $x$ variables and at least one $y$ variable, we get no reductions to 0 which means that the system could be semi-regular or $\gamma$-semi-regular.

		Notes pour moi: On a quelque chose de bizarre où si on prends la 2 de la proposition \ref{PropSemi}, on a pas semi-régularité voir rien de particulier. Mais si on prends la 1, donc oublier qu'on est sur $\mathbb{F}_2$, on a toujours pas de semi-regularité d'après la proposition, mais la 5 est vérifié avec la série generatrice, donc je soupçonne que la façon de laquelle sagemath calcule la série de hilbert est pas forcément correct, à vérifier avec des exemples de suites semi-regulière. Update: Il semblerait qu'avec des systèmes aléatoires, donc qui devrait être semi-régulier, sagemath contredit la 5 de \ref{PropSemi}. Donc il faut voir si la méthode .hilbert\_series est la bonne ou s'il faut utiliser autre chose. Update 2: Sage tronque la série de hilbert jusqu'au degré de régularité très certainement et ça correspond à la série génératrice mais seulement avec la formule en 1 de \ref{PropSemi} et non la 2 ce qui est pas normal vu qu'on est sur $\mathbb{F}_2$. Il est possible que ça soit la méthode de sage qui fait n'importe quoi quand on est sur $\mathbb{F}_2$. Après vérification, c'est bien ce que sagemath fait donc il prends pas en compte qu'on est sur $\mathbb{F}_2$.

		\subsection{Bilinear systems case}
		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2, overdetermined and can be semi-regular.

@@

		We will denote $n_x$ (resp. $n_y$) the number of variables in $x$ (resp. number of variables in $y$). In our case, $n_x = n_y = (m-1)/2$ and $m = k$. As we mentioned earlier, our polynomial systems can be semi-regular, so we get no reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems until the Macaulay matrix of degree $D_{reg}$.

@@

		 No need to say that it's useless to use the criterion as shown in \cite{FSS11} without adapting it to get an advantage.
		 This can be explained by the fact that the extended criterion computes reduced polynomials of degree $n_x$ and $n_y$ and in our case, $n_x$ and $n_y$ can be too big to compute. Maybe there could be a way to compute reduced polynomials of only degree 2 for our use case if we stop Macaulay matrix computations at degree 4 with our quadratic systems.

		 As the authors of \cite{FSS11} noted at the end of their article, generic bilinear systems are regular. They didn't show that for overdetermined systems but we saw that experimentally but it's yet to be proven.

		 Note pour moi-même, il faut maintenant essayer de réaliser une implémentation de ce critère en sage ou avec hpXbred \cite{BS23} et BeanPolE \cite{Bou24} pour voir comment ça se comporterait avec des petits systèmes et voir quel serait l'avantage ou le désavantage sachant que la supposition que toutes les lignes se réduisent à 0 est extrêment peu probable. Donc les lignes se réduisent pas à 0 quand on a les x et les y donc le paramètre v de hpXbred est strictement supérieur à $n_x$.
		 On peut ensuite utiliser XL \cite{CKPS00} avec l'implantation \cite{NNY17}

>>>>

\author{Paul Mekhail\\ \textbf{Advisor:} Sorina Ionica}

@@

		\begin{definition}
			Let $\prec$ be an admissible monomial ordering on a given ring $R$. Given a sequence of homogeneous (resp. affine) polynomials $\mathcal{F} = (f_1,\dots,f_m) \in R$, we associate to it the Macaulay matrix of degree $D$ (resp. $\leq D$), denoted by $\Mac{D}{m}{\mathcal{F}}$ (resp. $\Mac{\leq D}{m}{\mathcal{F}}$), and defined as follows: the columns of the matrix are indexed by the monomials of degree $D$ (resp. $\leq D$) sorted in decreasing order from the left to right using the defined monomial ordering. Each row of the matrix is labeled by a tag, also called signature, ($u, f_i$) where u is a monomial in $R$ and $f_i \in \mathcal{F}$ such that $deg(uf_i) = D$ (resp. $deg(uf_i) \leq D$), and the polynomial $uf_i$ is written as vector of coefficients of monomials on the row. We denote by $\Mact{D}{m}{F}$ (resp. $\Mact{\leq D}{m}{F}$) the row echelon form of $\Mac{D}{m}{\mathcal{F}}$ (resp. $\Mac{\leq D}{m}{\mathcal{F}}$) without swapping columns to preserve the monomial ordering.
		\end{definition}

@@

		The major "issue" of Gröbner basis algorithms like Bucherberger's \cite{Buc}, F4 \cite{F02} are the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
		reductions to 0 can cost a lot during the linear algebra process. F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.

@@

		\subsection{Hilbert series}
		Hilbert-Poincaré series (or just Hilbert series) is a very important tool in algebraic geometry to describe ideals, varieties and their dimensions.
		\begin{definition}[\cite{CLS}]
			Let $I$ be a homogeneous ideal in $R$. The affine Hilbert function of $I$ is the function on the nonnegative integers s defined by
			$$
				{}^a HF_{R/I}(s) = dim R_{s} / dim I_{s} = dim R_{s} - dim I_{s}.
			$$
		\end{definition}

@@

		Thus, Bardet studied a new concept called semi-regularity for overdetermined system to allow their study and analysis. The definition of a semi-regular sequence is given by Bardet \cite{Bardet04}.

@@

@@

		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $H(I) - 1$, where $I$ is the ideal generated by the polynomials of the system.

@@

		\subsection{Bilinear systems case}
		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2 and overdetermined.

@@

		We will denote $n_x$ (resp. $n_y$) the number of variables in $x$ (resp. in $y$). In our case, $n_x = n_y = (m-1)/2$ and $m = k$.

@@

	\section{Boolean systems}
		Sometimes, the finite field used is $\mathbb{F}_2$ because it is still supposedly hard to solve but is very easy to use and manipulate on computers. But in such small fields one can begin to use exhaustive search to obtain the solution to the system. The best algorithm known to the authors for exhaustive search for multivariate polynomial system $f_1,\dots,f_m$ over $\mathbb{F}_2[x_1,\dots,x_n]$ is \cite{BCC+10} with a complexity $\mathcal{O}(ln(n)2^n)$, what is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$. There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textbf{libFES}.

		The last type is a hybrid algorithms using gröbner basis algorithms and exhaustive search to find solutions. The most known and the best one is BooleanSolve hybrid by Bardet, Faugère, Salvy and Spaenlehauer \cite{BFSS13} where we guess a certain number of variables $k$ and apply F5 algorithm on the specialized systems. Under semi-regularity assumption, the algorithm finds the solution in $\mathcal{O}^(2^{0.792n})$ if $n = m$ and $k = 0.55n$. This is better than exhaustive search, but still, the constant factors are too big for it to be better for systems smaller with less than 200 variables.

		\begin{algorithm}
			\caption{The CrossBred algorithm}\label{alg:crossbred}
			\begin{algorithmic}[1]
				\Require Polynomials $\mathcal{F} = {f_1, \dots, f_m}$ of $n$ variables and integers $D,d,k$
				\Ensure  A solution for the system if it exists or nothing otherwise.
				\State Construct $\Mac{arg1}{arg2}{arg3}$
				\State Compute a basis $(v_1,\dots,v_r)$ if the left kernel of ...
				\State Construct the polynomials $p_1,\dots,p_r$ corresponding ti ...
				\For{$a = (a_{k+1},\dots,a_n) \in \mathbb{F}_{2}^{n-k}$}
				\State Evaluate the last $n-k$ variables in each $f_i \in \mathcal{F}$ at $(a_{k+1},\dots,a_n)$ and compute $\mathcal{F}*$
				\State Compute $Mac_{\leq d, m}(\mathcal{F}*)$
				\State Compute $\mathcal{F}'*$ as the partial evaluation of $F' = {p_1,\dots,p_r}$ at $(a_{k+1},\dots,a_n)$
				\State Consider the system $S*$ consisting of $Mac_{\leq d, m}(\mathcal{F}*) \cup \mathcal{F}'*$
				\If{$S*$ is consistent using XL for example}
				\State \textbf{Return} solution
				\EndIf
				\EndFor
			\end{algorithmic}
		\end{algorithm}

		This algorithm have shown it's effectiveness in practice with good implementations like Bouillaguet and Sauvage's implementation \cite {BS23} that broke records for solving boolean systems on the Fukuaka MQ challenge \cite{MQFUK}.

	\section{Experimental results}
		We tested if small instances of our problem are semi-regular or not using Proposition \ref{PropSemi} and they were not semi-regular. We also remarked that the bilinear criterion \ref{algo:crit} detected reductions to 0 using the original F5 criterion \ref{F5Crit} and Frobenius criterion for systems over $\mathbb{F}_2$ but also detected reductions to 0 on polynomials of degree $n_x$ or $n_y$ according to the algorithms \ref{alg:crit_calc} and \ref{alg:reduc}. In our case, we will use the Crossbred algorithm and we do not think that we will get to these degrees in the Macaulay matrices, so no need to use the bilinear criterion.

		Even though no concrete complexity analysis is available for the Crossbred algorithm because of it's wide parameters choice, there are some articles that give formal tools to analyze admissible parameters and try to estimate the complexity for these parameters. We choose to focus on the article \cite{VID24} to select valid parameters and estimate the complexity of the algorithm using them.

		\begin{theorem}
			Let $H_{m, n}^k(X,Y) = \sum_{d_1 \geq 0, d_2 \geq 0} h_{d1, d2, m}^k X^{d_1}Y^{d_2}$ be the bivariate series with coefficient which is the number of new "independent" polynomials generated during the pre-processing. This series is given by:
			$$
				H_{m, n}^k(X, Y) = \frac{1}{Y}\left((1+X)^{n-k} - \frac{(1+XY)^k(1+X)^{n-k}}{(1+X^2Y^2)^m}\right).
			$$
		\end{theorem}

		\begin{proposition}
			For fixed values of $m, n \text{ and } k$ the bivariate series $G^{k}_{m,n}(X, Y) = $ \\ $\sum_{d_1 \geq 0, d_2 \geq 0}(\sum_{d_{1}' \leq d_1, d_{2}' \geq d_2} g_{d_{1}',d_{2}',m}^k)X^{d_1}Y^{d_2}$ is given by the formula:
			$$
				G^{k}_{m,n}(X, Y) = - \frac{YH_{m, n}^k(X,Y) - H_{m, n}^k(X, 1)}{(1-X)(1-Y)}.
			$$

			the coefficient in front of $X^{d_1}Y^{d_2}$ is the number of polynomials generated during the pre-processing step for a given pair $(d_1, d_2)$.
		\end{proposition}

		\begin{theorem}
			Let $\mathcal{F} = {f_1,\dots,f_m}$ be a $\gamma$-strong semi-regular sequence of polynomials in $\mathbb{F}_2[x1,\dots,x_n]$ and consider $D < D_{reg}, d < d_{reg}(k).$ Then $k, D$ and $d$ are potentially admissible parameters for the Crossbred algorithm if the coefficient corresponding to $X^DY^d$ of the following bivariate series
			$$
			J_{m, n}^k(X, Y) = \frac{1}{(1-X)(1-Y)}\left(\frac{(1+X^{n-k})(1+XY)^k}{(1+X^2Y^2)^m} - \frac{(1+X)^n}{(1+X^2)^m} - \frac{(1+Y)^k}{(1+Y^2)^m}\right)
			$$
			is non-negative.
		\end{theorem}

		We wrote sage scripts that automated the search of admissible parameters using the bivariate series given in \cite{VID24}. We searched for different key sizes to have a rough estimate on the complexity.

		For the complexity estimate for the precomputation step we took for reference the complexity of block-lanczos and block-wiedemann algorithms which is $\mathcal{O}(n*nnz)$ where $n$ is the number of rows and $nnz$ is the number of non zero elements of the matrix.

		For the space complexity we thought that the best is to consider sparse data storage using CSR format where we store 2 arrays, one array is storing the number of non zero elements per row and the second array stores the columns of the non zero elements. No need to store the coefficients because they are always $1$. So the storage complexity is $\mathcal{O}(nb\_rows + nnz)$ bytes.
